Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-20 popl-rebuttal.md

- Local Context Independence: This assumption is used by the inference
  rule for `FOREACH`. The rule computes the transformer of a loop by
  lifting the transformer (${\sf F}$) of its body. In the process, it
  assumes that the effect of each loop iteration depends only on
  the loop variable $z$, but not on local ($\delta$) and global
  ($\Delta$) states (observe that ${\sf F}$ is applied to the same
  $\delta$ and $\Delta$ forall $z$). While independence on $\Delta$ is
  gauranteed by the stability condition (checked separately), the
  independence on the local state $\delta$ is simply assumed.

Reviewer C
-------------

- Dealing with heap state: It is indeed true that we have ignored heap
  memory to focus exclusively on database state in our
  formalization. Swamy et al. (PLDI'13), however, have proposed the
  Dijkstra monad to reason about heap memory monadically.  As part of
  future work, we intend to explore the composition of the DB and
  Dijkstra monads to enable reasoning over both heap and database
  state.

Shepard
----------
- The primary concern raised by reviewer B is that the fragment of FOL
  in which the translation algorithm (Fig. 10, Sec.5) encodes set
  expressions is not EPR, hence its decidability is doubtful. In
  particular, the concern was that an existential quantifier follows a
  universal quantifier in the logical encoding of bind (`>>=`), and
  this pattern is not admitted by EPR. In the rebuttal phase we
  addressed the concern by pointing out that the fragment, while not
  EPR as claimed, is nonetheless decidable since it is subsumed by the
  GKS fragment, which is known to be decidable. However, as the
  reviewer rightly points out, there are no known results on the
  combination of GKS and Simple Linear Arithmetic (SLA) to support
  theorem 5.3, so SLA invariants, which are common in databases, might
  make the encoding undecidable. While this concern can be addressed
  by considering machine (finite-bit) arithmetic instead of SLA, we
  decided to take a long, hard look at the encoding algorithm and see
  if it can be simplified. In this we were successful as we were able
  to modify our encoding to eliminate the existential altogether. Our
  modifications are based on the observation that the semantics of
  bind, which were previously written using two implications one of
  which involves an existential, can be equivalently written using a
  single bi-implication that only involves universal quantifiers. A
  rigorous proof to this effect can be found in the techreport, while
  an automatic Z3-assisted proof can be found
  [here](https://rise4fun.com/Z3/G1a). As a consequence of this
  change, the fragment of encoding is back to being EPR, hence the
  Theorems 5.2 and 5.3 continue to hold.

### 2017-10-20 inference.tex
edits

@@

@@

The correspondence between the inference rules and the RG judgment of
\S~\ref{sec:reasoning} is stated thus:
\begin{theorem}
\label{thm:inference-sound}
  \emph{Forall} i,$R$,$I$,$c$,$\F$, if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}
\noindent where the set expression $\F(\emptyset,\stg)$ has a natural
interpretation in the theory of sets.

\begin{proof}
  We only provide a brief proof sketch here\footnote{The complete proof can be found in
the supplementary material.}. We prove a slightly
  stronger statement :

  \emph{Forall} i,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c
  \elabsto \F$, then:

  \begin{smathpar}
    \begin{array}{c}
    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
    I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
    \end{array}
  \end{smathpar}

  We use structural induction on $c$. Assume
  that $P \Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$
  and $Q \Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For
  each type of $c$, our strategy is to show the premises of the
  corresponding RG-rule. The base cases correspond to the INSERT,
  UPDATE and DELETE statements. In all these cases, the form of $F$
  directly corresponds to the condition linking the pre-condition $P$
  and post-condition $Q$ in the premises of the RG-rules. If
  $\stable(\R, F)$, then $\stabilize{F} = F$ and hence the proof
  directly follows. Otherwise, $\stabilize{F} = \lambda (\stl,
  \stg).\existsl(\stg', I(\stg'), F(\stl, \stg'))$. Since the
  pre-condition $P(\stl, \stg)$ asserts $I(\stg)$, any state which
  obeys $P$ provides the existential $\stg'$ in the above set
  expression, and hence $F$ applied on that state matches the
  condition linking $P$ and $Q$. The stability of $P$ with respect to
  $\R$ follows directly from the stability of $I$.

  The proofs for SELECT, SEQ and IF-THEN-ELSE directly use the inductive
  hypothesis to infer the RG-judgements present in the premises of
  their corresponding RG-rules. The interesting case is the FOREACH
  statement, for which we use the loop invariant $\psi(\stl, \stg)
  \Leftrightarrow \stl = s \cup (y \bind (\lambda z.~\F(s, \stg)))$,
  (where assuming that $c$ is the body of the loop, $c \elabsto F$) to
  prove all the premises of \rulelabel{RG-Foreach}. Using the same
  notation as the rule $\rulelabel{RG-Foreach}$, $y$ refers to the
  records already processed in previous iterations of the loop, while
  $z$ refers to the record being processed in the current iteration.
  At the beginning of the loop $[\phi/y]\psi(\stl, \stg)$ just reduces
  to $\stl = s$ which is implied by the pre-condition $P$. From the
  inductive hypothesis, we can infer that each iteration corresponds
  to the application of $F$. Since all iterations are assumed to be
  independent of each other, and $z$ is bound to a record in $x$ for
  each iteration, we can infer that at the end of every iteration the
  loop invariant $[y \cup \{z\}/y]\psi$ will be satisfied. 
\end{proof}

@@

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts a
an $x:T$ and returns a quantifier-free proposition that evaluates to
true ($\top$) if and only if $x$ is present in the set. Alternatively,
translation may also encode the set as a function in the logic itself,
in which case a quantified proposition constraining the function is
also generated. For instance, consider the set $\{1,2\}$. The
predicate describing the set can be encoded as the function $\lambda
\v. \v=1 \vee \v=2$, with no further constraints (i.e., with $\top$
constraint), or it can be encoded as the function $\lambda \v. g(\v)$
with an associated constraint ($\phi\in\Prop^1$) defining the
uninterpreted boolean functon $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted functions used in $G$.

Due to the presence of bind ($\bind$) in $\SL$, a set expression $s$
may contain free variables introduced by an enclosing binder. For
instance, consider the $\SL$ expression $s_1\bind(\lambda x. \{ y\,|\,
y=x+1\})$, where $s_1$ is an integer set (expression). The
subexpression $\{ y\,|\, y=x+1\}$ (call it $s_2$) contains $x$ as a
free variable. In such cases, the predicate associated with the
subexpression should also be indexed by its free variables, so that a
unique set exists for each instantiation of the free variables. For
instance, the predicate ($\G$) associated with subexpression from the
above example should be $\lambda\v_1.\lambda\v_2.~\v_2 = \v_1 + 1$, so
that the set $\G\; x_1$ is different from the set $\G\; x_2$ for
distinct $x_1,x_2 \in s_1$. Intuitively, the bind expression
$s_1\bind(\lambda x. \{ y\,|\, y=x+1\})$ denotes the set
$\bigcup\limits_{x\in s_1}\G\;x$.

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
($\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similarly.

The first-order encoding of a bind expression describes the semantics
of set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain $(\phi_1,\G_1)$
and $(\phi_2,\G_2)$, respectively. Since $s_2$ is under a new binder
that binds $x$, free variable sequence of $s_2$ is $\nubar,x$. In the
interest of hygiene, we substitute a fresh $a$ for $x$, making the
sequence $\nubar,a$. The set $s$ is encoded as a a new uninterpreted
predicate $g$ indexed by $s$'s free variables ($\nubar$). Since the
set denoted by $g$ is the result of the bind $s_1 \bind \lambda
x.s_2$, a first-order constraint defining the bind operation (as
described above) is generated. The constraint relates the predicates
$\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to the
uninterpreted predicate $g$ that represents $s$. The constraint is
assigned a name ($\pi$) so that it can be enabled only for the choice
of free variables ($\nubar$) for which $g$ needs to be defined;
observe that the lambda returned by $\mssemof{\nubar}{s_1 \bind
\lambda x.s_2}$ asserts $\pi(\vbar)$ only in conjunction with
$g(\vbar,r)$, for some $r$.

The first-order encoding of the $\existsl(\stg,\phi,s)$ expression
eliminates the existential by Skolemizing it. Skolemizing is the
process of substituting an existentially bound $x$ in
$\phi_x\in\Prop^1$ with $f(\nubar)$, where $f$ is a fresh
uninterpreted function, and $\nubar$ are the free variables in
$\phi_x$ bound by enclosing universal quantifiers. Due to the
decidability restrictions (Sec.~\ref{sec:decidability}), the only
uninterpreted functions we admit in our logic are boolean (i.e.,
predicates/relations). Consequently, Skolemizing an existentially
bound $\stg$ with $f(\nubar)$ is not really an option in our case.
Instead, we define $f$ as an uninterpreted relation that behaves as a
partial function:
\begin{smathpar}
  \begin{array}{c}
    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
    \Rightarrow a = b
  \end{array}
\end{smathpar}
We then replace the existentially bound $\stg$ with a new universally
bound $a$ in $\phi$ and $s$, and define the set $[a/\stg]s$ under the
condition that $f(\nubar,a) \conj [a/\stg]\phi$ holds. Since the
existence of a $\stg$ that satisfies $\phi$ is axiomatized tp
define $s$, if the aforementioned condition does not hold, i.e., if
there does exist such a $\stg$, we let the context derive $\bot$.

\noindent {\bf Example} Let us reconsider the  TPC-C \C{new\_order}
transaction from Sec.~\ref{sec:motivation}. Recall that the state
transformer ($\T$) for the foreach loop shown in
Fig.~\ref{fig:foreach_code} is (\C{k1}, \C{k2}, and \C{k3} are
constants):
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
Where:
\begin{smathpar}
  \begin{array}{lcl}
    \F_U & = & \lambda(\stl,\stg).~ \stg \bind(\lambda s. 
                \itel{{\sf table}(s) = \C{Stock} \conj 
                  s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
         &   & \hspace*{0.9in}}
                     {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                                 \C{s\_d\_id}=s.\C{s\_d\_id};\,
                                 \C{s\_qty} = \C{k1}\rangle \}\\
         &   & \hspace*{0.9in}} {\emptyset}) \\
    \F_I & = & \lambda(\stl,\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,
                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}\\
  \end{array}
\end{smathpar}
For any $\stl$ and $\stg$, $\F_U(\stl,\stg)$ and $\F_I(\stl,\stg)$ are
expressions in $\SL$, so can be translated to FOL by the encoding
algorithm in Fig.~\ref{fig:logic}. Since the iteration variable
\C{item\_req} occurs free in these expressions, the appropriate
application of the encoding algorithm is
$\mssemof{\C{item\_req}}{\F_U(\stl,\stg)}$ and
$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which result in
$(\phi_U,\G_U)$ and $(\phi_I,\G_I)$, respectively, where $\phi_U$,
$\phi_I$, $\G_U$, $\G_I$ are as shown below:
\begin{smathpar}
  \begin{array}{lcl}
    \phi_U & = & \forall \C{item\_req}.\forall s. \forall s'.~  \\
    & & \hspace*{0.1in}\pi(\C{item\_req}) \Leftrightarrow(s \in \stg) \conj 
        (\itec{{\sf table}(s)=\C{Stock} \wedge 
              s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
    & & \hspace*{1.45in}}
              {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                            \C{s\_d\_id}=s.\C{s\_d\_id};\,
                            \C{s\_qty} = \C{k1}\rangle\\
    & & \hspace*{1.45in}} {\bot}) \\
    & & \hspace*{1.2in} \Leftrightarrow g(\C{item\_req},s')\\
    \G_U & = & \lambda (\C{item\_req},r).~ \pi(\C{item\_req}) 
        \wedge g(\C{item\_req},r)\\
    \phi_I & = & \top \\ 
    \G_U & = & \lambda (\C{item\_req},r).~ 
        r = \langle \C{ol\_o\_id}=\C{k2},\,
                    \C{ol\_d\_id}=\C{k3},\, 
                    \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                    \C{ol\_qty}=\C{item\_req.ol\_qty}
             \rangle\\
  \end{array}
\end{smathpar}

>>>>

\begin{center}

@@

\end{center}

@@

We can now establish a correspondence between the inference rules given
above and the RG judgment of \S\ref{sec:reasoning}:
\begin{theorem}
  \label{thm:inference-sound}
  Let the set expression $\F(\emptyset,\stg)$ have a natural interpretation
  in the theory of sets.  Then, for all $i$,$R$,$I$,$c$,$\F$,
  if $\stable(\R,I)$ and $c \elabsto \F$,
  then:\\\vspace*{-0.2cm}
  \begin{smathpar}
  \begin{array}{c}
  \R \vdash \hoare{\lambda(\stl,\stg).~\stl=\emptyset \conj
  I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(\emptyset,\stg)}
  \end{array}
  \end{smathpar}
\end{theorem}

\paragraph{{\sc Proof Sketch.}} We prove a slightly stronger statement :
For all  $i$,$R$,$I$,$c$,$\F$,$s$ if $\stable(\R,I)$ and $c \elabsto \F$, then:
  \begin{smathpar}
    \begin{array}{c}
    \R \vdash \hoare{\lambda(\stl,\stg).~\stl=s \conj
    I(\stg)}{c}{\lambda(\stl,\stg).\stl = \F(s,\stg)}
    \end{array}
  \end{smathpar}
The proof follows by structural induction on $c$. Assume that $P
\Leftrightarrow \lambda(\stl,\stg).~\stl=s \conj I(\stg)$ and $Q
\Leftrightarrow \lambda(\stl,\stg).\stl = \F(s,\stg)$. For each type
of $c$, our strategy is to show the premises of the corresponding
RG-rule. The base cases correspond to \C{INSERT}, \C{UPDATE} and \C{DELETE}
statements. In all these cases, the form of $\F$ directly corresponds
to the condition linking the pre-condition $P$ and post-condition $Q$
in the premises of the RG-rules. If $\stable(\R, \F)$, then
$\stabilize{\F} = \F$ and hence the proof directly follows. Otherwise,
$\stabilize{\F} = \lambda (\stl, \stg).\existsl(\stg', I(\stg'),
\F(\stl, \stg'))$. Since the pre-condition $P(\stl, \stg)$ asserts
$I(\stg)$, any state which obeys $P$ provides the existential $\stg'$
in the above set expression, and hence $\F$ applied on that state
matches the condition linking $P$ and $Q$. The stability of $P$ with
respect to $\R$ follows directly from the stability of $I$.

The proofs for \C{SELECT}, sequencing, and conditionals use use the
inductive hypothesis to infer the RG-judgements present in the
premises of their corresponding RG-rules. The interesting case is the
\C{FOREACH} statement, for which we use the loop invariant
$\psi(\stl, \stg) \Leftrightarrow \stl = s \cup (y \bind (\lambda
z.~\F(s, \stg)))$, (where assuming that $c$ is the body of the loop,
$c \elabsto \F$) to prove all the premises of
\rulelabel{RG-Foreach}. Using the same notation as the rule
$\rulelabel{RG-Foreach}$, $y$ refers to the records already processed
in previous iterations of the loop, while $z$ refers to the record
being processed in the current iteration.  At the beginning of the
loop $[\phi/y]\psi(\stl, \stg)$ just reduces to $\stl = s$ which is
implied by the pre-condition $P$. From the inductive hypothesis, we
can infer that each iteration corresponds to the application of
$F$. Since all iterations are assumed to be independent of each other,
and $z$ is bound to a record in $x$ for each iteration, we conclude
that at the end of every iteration, the loop invariant $[y \cup
  \{z\}/y]\psi$ will be satisfied.

@@

The algorithm ($\mssemof{\cdot}{\cdot}$) shown in Fig.~\ref{fig:logic}
translates an $\SL$ expression ($s$) to FOL. The translation is based
on encoding a set of element type $T$ as a unary predicate on $T$.
The predicate is represented as a meta function that accepts an
$x:T$ and returns a quantifier-free proposition that evaluates to
true ($\top$) if and only if $x$ is present in the set. Alternatively,
translation may also encode the set as a function in the logic itself,
in which case a quantified proposition constraining the function is
also generated. For instance, consider the set $\{1,2\}$. The
predicate describing the set can be encoded as the function $\lambda
\v. \v=1 \vee \v=2$, with no further constraints, or it can be encoded as the function $\lambda \v. g(\v)$
with an associated constraint ($\phi\in\Prop^1$) defining the
uninterpreted boolean functon $g$:
\begin{smathpar}
  \begin{array}{lcl}
    \phi & = & \forall \nu.~g(\nu) \Leftrightarrow \nu=1 \vee \nu=2
  \end{array}
\end{smathpar}
The translation adopts one or the other approach, depending on the
need. For uniformity, we consider the encoding of a set as pair
($\phi$,$\G$), where $\G$ is a meta function, and $\phi$ is a FOL
formula constraining any uninterpreted functions used in $G$.

Due to the presence of bind ($\bind$) in $\SL$, a set expression $s$
may contain free variables introduced by an enclosing binder. For
instance, consider the $\SL$ expression $s_1\bind(\lambda x. \{ y\,|\,
y=x+1\})$, where $s_1$ is an integer set (expression). The
subexpression $\{ y\,|\, y=x+1\}$ (call it $s_2$) contains $x$ as a
free variable. In such cases, the predicate associated with the
subexpression should also be indexed by its free variables, so that a
unique set exists for each instantiation of the free variables. Thus,
the predicate ($\G$) associated with the subexpression from the
above example should be $\lambda\v_1.\lambda\v_2.~\v_2 = \v_1 + 1$, so
that the set $\G\; x_1$ is different from the set $\G\; x_2$ for
distinct $x_1,x_2 \in s_1$. Intuitively, the bind expression
$s_1\bind(\lambda x. \{ y\,|\, y=x+1\})$ denotes the set
$\bigcup\limits_{x\in s_1}\G\;x$.

The translation algorithm (Fig.~\ref{fig:logic}) takes free variables
into account. Given a set expression $s\in\SL$, whose (possible) free
variables are $\nubar$ in the order of their introduction (top-most
binder first), $\mssemof{\nubar}{s}$ returns the encoding of $s$ as
$(\phi,\G)$.  The meta function $\G$ is a predicate indexed by the
(possible) free variables of $s$, thus, its arity is $|\nubar| + 1$.
Note that $\nubar$ is only a sequence of variables introduced by the
enclosing binders of $s$, and not all may actually occur free in $s$.
Nonetheless, its predicate $\G$ is always indexed by $|\nubar|$ free
variables for uniformity. The translation encodes database state as an
uninterpreted sort. Considering that the state is actually a set of
records, we define an uninterpreted relation $\in$ to relate records
and states. Thus, a variable set expression $\stg$ denoting a database
state is encoded as the predicate $\lambda (\vbar,r).~r\in\stg$, where
$|\vbar|=|\nubar|$ (predicates are uncurried for simplicity; $\vbar$
is a comma-separated sequence; $r\not\in\SL$ is a special variable).
The constraints associated with the encoding of a state are trivial
(denoted $\top$). The set literal expression $\{x\,|\, \varphi\}$ is encoded
straightforwardly. The conditional set expression is encoded as an
if-then-else predicate in FOL, where the predicates on true and false
branches are computed from the set subexpressions $s_1$ and $s_2$,
respectively. The conjunction of constraints $\phi_1$ and $\phi_2$,
from $\mssemof{\nubar}{s_1}$ and $\mssemof{\nubar}{s_2}$ (resp.), is
propagated upwards as the constraint of the conditional expression.
Set union expression is encoded similarly.

The first-order encoding of a bind expression describes the semantics
of the set monad's bind operator in FOL. Let $s_1$ be a set, and let $f =
\lambda x.~s_2$ be a function that maps each variable in $s_1$ to a
new set. Then, $s = s_1 \bind f$ if and only if forall $x$, $y$, if
$x\in s_1$ and $y \in f(x)$, then  $y \in s$, and vice versa. The
translation first encodes $s_1$ and $s_2$ to obtain $(\phi_1,\G_1)$
and $(\phi_2,\G_2)$, respectively. Since $s_2$ is under a new binder
that binds $x$, the free variable sequence of $s_2$ is $\nubar,x$. In the
interest of hygiene, we substitute a fresh $a$ for $x$, making the
sequence $\nubar,a$. The set $s$ is encoded as a new uninterpreted
predicate $g$ indexed by $s$'s free variables ($\nubar$). Since the
set denoted by $g$ is the result of the bind $s_1 \bind \lambda
x.s_2$, a first-order constraint defining the bind operation (as
described above) is generated. The constraint relates the predicates
$\G_1$ and $\G_2$, representing $s_1$ and $s_2$ (resp.), to the
uninterpreted predicate $g$ that represents $s$. The constraint is
assigned a name ($\pi$) so that it can be enabled only for the choice
of free variables ($\nubar$) for which $g$ needs to be defined;
observe that the lambda expression returned by $\mssemof{\nubar}{s_1 \bind
\lambda x.s_2}$ asserts $\pi(\vbar)$ only in conjunction with
$g(\vbar,r)$, for some $r$.

The first-order encoding of the $\existsl(\stg,\phi,s)$ expression
eliminates the existential by Skolemizing it. Skolemizing is the
process of substituting an existentially bound $x$ in
$\phi_x\in\Prop^1$ with $f(\nubar)$, where $f$ is a fresh
uninterpreted function, and $\nubar$ are the free variables in
$\phi_x$ bound by enclosing universal quantifiers. Due to the
decidability restrictions (Sec.~\ref{sec:decidability}), the only
uninterpreted functions we admit in our logic are boolean (i.e.,
predicates/relations). Consequently, Skolemizing an existentially
bound $\stg$ with $f(\nubar)$ is not really an option in our case.
Instead, we define $f$ as an uninterpreted relation that behaves as a
partial function:
\begin{smathpar}
  \begin{array}{c}
    \forall \nubar.\forall a.\forall b.~f(\nubar,a) \wedge f(\nubar,b)
    \Rightarrow a = b
  \end{array}
\end{smathpar}
We then replace the existentially bound $\stg$ with a new universally
bound $a$ in $\phi$ and $s$, and define the set $[a/\stg]s$ under the
condition that $f(\nubar,a) \conj [a/\stg]\phi$ holds. Since the
existence of a $\stg$ that satisfies $\phi$ is axiomatized to
define $s$, if the aforementioned condition does not hold, i.e., if
there does exist such a $\stg$, we let the context derive $\bot$.

\noindent {\bf Example} Let us reconsider the  TPC-C \C{new\_order}
transaction from Sec.~\ref{sec:motivation}. Recall that the state
transformer for the \C{foreach} loop shown in
Fig.~\ref{fig:foreach_code} is (\C{k1}, \C{k2}, and \C{k3} are
constants):
\begin{smathpar}
\begin{array}{l}
  \lambda(\stl,\stg).~ \stl \cup \C{item\_reqs}\bind
      (\lambda\C{item\_req}.~ \F_U(\stl,\stg) \cup \F_I(\stl,\stg))
\end{array}
\end{smathpar}
where:
\begin{smathpar}
  \begin{array}{lcl}
    \F_U & = & \lambda(\stl,\stg).~ \stg \bind(\lambda s. 
                \itel{{\sf table}(s) = \C{Stock} \conj 
                  s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
         &   & \hspace*{0.9in}}
                     {\{ \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                                 \C{s\_d\_id}=s.\C{s\_d\_id};\,
                                 \C{s\_qty} = \C{k1}\rangle \}\\
         &   & \hspace*{0.9in}} {\emptyset}) \\
    \F_I & = & \lambda(\stl,\stg).~ \{\langle\C{ol\_o\_id}=\C{k2},\,
                 \C{ol\_d\_id}=\C{k3},\, \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                 \C{ol\_qty}=\C{item\_req.ol\_qty}\rangle\}\\
  \end{array}
\end{smathpar}
For any $\stl$ and $\stg$, $\F_U(\stl,\stg)$ and $\F_I(\stl,\stg)$ are
expressions in $\SL$, so can be translated to FOL by the encoding
algorithm in Fig.~\ref{fig:logic}. Since the iteration variable
\C{item\_req} occurs free in these expressions, the appropriate
application of the encoding algorithm is
$\mssemof{\C{item\_req}}{\F_U(\stl,\stg)}$ and
$\mssemof{\C{item\_req}}{\F_I(\stl,\stg)}$, which results in
$(\phi_U,\G_U)$ and $(\phi_I,\G_I)$, respectively, where $\phi_U$,
$\phi_I$, $\G_U$, $\G_I$ are as shown below:
\begin{smathpar}
  \begin{array}{lcl}
    \phi_U & = & \forall \C{item\_req}.\forall s. \forall s'.~  \\
    & & \hspace*{0.1in}\pi(\C{item\_req}) \Leftrightarrow(s \in \stg) \conj 
        (\itec{{\sf table}(s)=\C{Stock} \wedge 
              s.\C{s\_i\_id} = \C{item\_req.ol\_i\_id}\\
    & & \hspace*{1.45in}}
              {s' = \langle \C{s\_i\_id}=s.\C{s\_i\_id};\, 
                            \C{s\_d\_id}=s.\C{s\_d\_id};\,
                            \C{s\_qty} = \C{k1}\rangle\\
    & & \hspace*{1.45in}} {\bot}) \\
    & & \hspace*{1.2in} \Leftrightarrow g(\C{item\_req},s')\\
    \G_U & = & \lambda (\C{item\_req},r).~ \pi(\C{item\_req}) 
        \wedge g(\C{item\_req},r)\\
    \phi_I & = & \top \\ 
    \G_U & = & \lambda (\C{item\_req},r).~ 
        r = \langle \C{ol\_o\_id}=\C{k2},\,
                    \C{ol\_d\_id}=\C{k3},\, 
                    \C{ol\_i\_id}=\C{item\_req.ol\_i\_id},\, 
                    \C{ol\_qty}=\C{item\_req.ol\_qty}
             \rangle\\
  \end{array}
\end{smathpar}

### 2017-10-20 opsem.tex
edits

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($[\bar{f}=\bar{e}]$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$), 
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}).

>>>>

Fig.~\ref{fig:txnimp} shows the syntax and small-step semantics of
\txnimp, a core language that we use to formalize our intuitions about
reasoning under weak isolation. Variables ($x$), integer and Boolean
constants ($k$), records ($r$) of named constants, sets ($s$) of such
records, arithmetic and boolean expressions ($e_1 \odot e_2$), and
record expressions ($[\bar{f}=\bar{e}]$) constitute the syntactic
class of expressions ($e$). Commands ($c$) include $\cskip$,
conditional statements, \C{LET} constructs to bind names, \C{FOREACH}
loops, SQL statements, their sequential composition ($c_1;c_2$),
transactions ($\ctxn{i}{\I}{c}$) and their parallel composition
($c_1\,||\,c_2$). Each transaction is assumed to have a unique
identifier $i$, and executes at the top-level; our semantics does not
support nested transactions. The $\I$ in the \C{TXN} block syntax is
the transaction's isolation specification, whose purpose is explained
below.  Certain terms that only appear at run-time are also present in
$c$.  These include a {\sf txn} block tagged with sets ($\stl$ and
$\stg$) of records representing local and global database state, and a
runtime {\sf foreach} expression that keeps track of the set ($s_1$)
of items already iterated, and the set ($s_2$) of items yet to be
iterated. Note that the surface-level syntax of the \C{FOREACH}
command shown here is slightly different from the one used in previous
sections; its higher-order function has two arguments, $y$ and $z$,
which are invoked (during the reduction) with the set of
already-iterated items, and the current item, respectively. This form
of \C{FOREACH} lends itself to inductive reasoning that will be useful
for verification (Sec.~\ref{sec:reasoning}).  Our language ensures
that all effectful actions are encapsulated within database commands,
and that all shared state among processes are only manipulated via
transactions and its supported operations.  In particular, we do not
consider programs in which objects resident on e.g., the OCaml heap
are concurrently manipulated by OCaml expressions as well as database
actions.
